### 2016-06-29 publications/Journal/2-background.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\vspace{1cm}
\subsection{Conservative Synchronization}
The first synchronization protocol we introduce is \textit{conservative synchronization}~\cite{FujimotoBook}.
In conservative synchronization, a node progresses independent of all other nodes, up to the point in time where it can guarantee that no causality errors happen.
When simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors occur.
In practice, this means that all nodes are aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
Deadlocks can occur due to a dependency cycle of models.
Multiple algorithms are defined in the literature to handle both the core protocol, as well as resolution schemes to handle or avoid the deadlocks~\cite{FujimotoBook}.

The main advantage of conservative synchronization is its low overhead if the lookahead is high.
Each node then simulates in parallel, and sporadically notifies other nodes about its local simulation time.
The disadvantage, however, is that the amount of parallelism is explicitly limited by the lookahead.
If a node can influence another (almost) instantaneously, no matter how rarely it occurs, the amount of parallelism is severely reduced.
The user is required to define the lookahead, using knowledge about the model's behaviour.
Defining lookahead is not always a trivial task if there is no detailed knowledge of the model.
Even slight changes in the model can change to the lookahead, and can therefore have a significant influence on simulation performance.

@@

Rolling back simulation time requires the node to store past model states, such that they can be restored later.
All incoming and outgoing events need to be stored as well.
Incoming events are injected again after a rollback, when their time has been reached again.
Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
In practice, a single causality error can have significant repercussions.

Further changes are required for unrecoverable operations, such as I/O and memory management.
These are only executed after the lower bound of all simulation times, called \textit{Global Virtual Time} (GVT), has progressed beyond their execution time.

The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states are stored up to a point where it can be considered \textit{irreversible}.
Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.

>>>>

\subsection{Conservative Synchronization}
The first synchronization protocol we introduce is \textit{conservative synchronization}~\cite{FujimotoBook}.
In conservative synchronization, a node progresses independent of all other nodes, up to the point in time where it can guarantee that no causality errors happen.
When simulation reaches this point, the node blocks until it can guarantee a new time until which no causality errors occur.
In practice, this means that all nodes are aware of the current simulation time of all other nodes, and the time it takes an event to propagate (called \textit{lookahead}).
Deadlocks can occur due to a dependency cycle of models.
Multiple algorithms are defined in the literature to handle both the core protocol, as well as resolution schemes to handle or avoid the deadlocks~\cite{FujimotoBook}.

The main advantage of conservative synchronization is its low overhead if the lookahead is high.
Each node then simulates in parallel, and sporadically notifies other nodes about its local simulation time.
The disadvantage, however, is that the amount of parallelism is explicitly limited by the lookahead.
If a node can influence another (almost) instantaneously, no matter how rarely it occurs, the amount of parallelism is severely reduced.
The user is required to define the lookahead, using knowledge about the model's behaviour.
Defining lookahead is far from a trivial task if there is no detailed knowledge of the model.
Even slight changes in the model or its allocation can change the lookahead, and can therefore have a significant influence on simulation performance.

@@

Rolling back simulation time requires the node to store past model states, such that they can be restored later.
All incoming and outgoing events need to be stored as well.
Incoming events are injected again after a rollback, when their time has been reached again.
Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
In practice, a single causality error can have significant repercussions on performance.

Further changes are required for unrecoverable operations, such as I/O and memory management.
These are only executed after the lower bound of all simulation times, called \textit{Global Virtual Time} (GVT)~\cite{FujimotoBook}, has progressed beyond their execution time.

The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
Apart from overhead in CPU time, a significant memory overhead is present: intermediate states are stored up to a point where it can be considered \textit{irreversible}.
Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.
Instead of depending on the theoretically defined safe lookahead, performance is related to the actually perceived lookahead.

### 2016-06-29 publications/Journal/3-features.tex

Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.

>>>>

Defining a lookahead function is therefore recommended in combination with conservative synchronization, but is not a necessity, as a default $\epsilon$ (\textit{i.e.}, the smallest representable timestep) is used otherwise.

### 2016-06-29 publications/Journal/memory.tex

\subsubsection{Remarks}
Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
Due to our motivation for memory usage analysis, we will only measure peak allocation.
Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.

>>>>

\subsubsection{Remarks}
Both dxex and adevs use \texttt{tcmalloc} as memory allocator, allowing for thread-local allocation.
Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
\texttt{tcmalloc} only gradually releases memory back to the OS, whereas our pools will not do so at all.
Due to our motivation for memory usage analysis, we will only measure peak allocation in maximum resident set size as reported by the OS.

### 2016-06-29 publications/Journal/models.tex

\subsection{Benchmarks}
We use three different benchmarks, which cover different aspects of the simulation kernel.

First, the \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
A single generator pushes events into the queue, which are consumed by the processors after a fixed or random delay.
It takes two parameters: width and depth, which determine the width and depth of the hierarchy.
This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
An example for width and depth 2 is shown in Figure~\ref{fig:queue_model}.

Second, the \textit{PHOLD} model, presented by~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
Each atomic model is directly connected to every other atomic model.
After a random delay, an atomic model sends out an event to a randomly selected output port.
Output port selection happens in two phases: first it is decided whether the event should be sent to an atomic model at the same node.
Afterwards, a uniform selection is made between the remaining ports.
The model takes two parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes, and the percentage of priority events. Priority events are events generated in a very short time after the previous event.
This benchmark shows how the simulation kernel behaves in the presence of many local or remote events and when the average time advance is larger than the lookahead in conservative simulation.
An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.

Third, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
The model takes one parameter: the number of models.
This benchmark investigates the complexity of event routing, and how the simulation kernel handles many simultaneous events.
An example for four models is shown in Figure~\ref{fig:interconnect_model}.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_model_fixed.pdf}
	\caption{Queue model for depth and width 2.}
	\label{fig:queue_model}

	\vspace{\betweenmodels}

	\includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
	\caption{Interconnect model for three models.}
	\label{fig:interconnect_model}

	\vspace{\betweenmodels}

	\includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
	\caption{PHOLD model for three models.}
	\label{fig:PHOLD_model}
\end{figure}

>>>>

\subsection{Benchmark Models}
We use three different benchmark models, covering different aspects of the simulation kernel.

First, the \textit{Queue} model, based on the \textit{HI} model of DEVStone~\cite{DEVStone}, creates a chain of hierarchically nested atomic \textsf{DEVS} models.
A single generator pushes events into the queue, which get consumed by the processors after a fixed or random delay.
It takes two parameters: the width and depth of the hierarchy.
This benchmark shows how the complexity of the simulation kernel behaves for an increasing amount of atomic models, and an increasingly deep hierarchy.
An example for width and depth equal to 2 is shown in Figure~\ref{fig:queue_model}.

Second, the \textit{PHOLD} model~\cite{PHOLD}, creates $n$ atomic models, where each model has exactly $n-1$ output ports.
Each atomic model is directly connected to every other atomic model.
After a random delay, an atomic model sends out an event to a randomly selected output port.
Output port selection happens in two phases: first it is decided whether the event should be sent within the same node, or outside of the node.
Afterwards, a uniform selection is made between the remaining models.
The model takes two parameter: the percentage of remote events, which determines the fraction of messages routed to other nodes, and the percentage of priority events. Priority events are events generated in a very short time after the previous event.
This benchmark shows how the simulation kernel behaves in the presence of many local or remote events, in combination with a varying percentage of high-priority event.s
An example for four models, split over two nodes, is shown in Figure~\ref{fig:PHOLD_model}.

Third, the \textit{Interconnect} model, a merge of PHOLD~\cite{PHOLD} and the \textit{HI} model of DEVStone~\cite{DEVStone}, creates $n$ atomic models, where each model has exactly one output port.
Similar to PHOLD, all models are connected to one another, but all through the same port: every model receives each generated event.
The model takes one parameter: the number of models.
This benchmark shows the complexity of event creation, event routing, and simultaneous event handling.
An example for four models is shown in Figure~\ref{fig:interconnect_model}.

\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_model_fixed.pdf}
	\caption{Queue model for depth and width 2.}
	\label{fig:queue_model}
\end{figure}

\begin{figure}
    \center
	\includegraphics[width=\modelfraction\columnwidth]{fig/interconnect_model.pdf}
	\caption{Interconnect model for three models.}
	\label{fig:interconnect_model}
\end{figure}

\begin{figure}
    \center
	\includegraphics[width=\modelfraction\columnwidth]{fig/phold_model.pdf}
	\caption{PHOLD model for three models.}
	\label{fig:PHOLD_model}
\end{figure}

### 2016-06-29 publications/Journal/parallel-performance.tex

We slightly modified the PHOLD benchmark, to include high-priority events.
Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
This situation closely mimics the case made in the comparison between both synchronization algorithms by~\cite{FujimotoBook}.

Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
When many high-priority events occur, optimistic synchronization is penalized due to its mindless progression of simulation, which frequently needs to be rolled back.
Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.

>>>>

We slightly modified the PHOLD benchmark to include high-priority events.
Contrary to normal events, high-priority events happen almost instantaneously, restricting lookahead to a very small value.
Even when normal events occur most often, conservative synchronization always blocks until it can make guarantees.
Optimistic synchronization, however, simply goes forward in simulation time and rolls back when these high-priority events happen.
This situation closely mimics the model typically used for comparing conservative and optimistic synchronization~\cite{FujimotoBook}.

Figure~\ref{fig:phold_priority} shows how simulation performance is influenced by the fraction of these high-priority events.
If barely any high-priority events occur, conservative synchronization is penalized due to its excessive blocking, which often turned out to be unnecessary.
When many high-priority events occur, optimistic synchronization is penalized due to its unconditional progression of simulation, which frequently needs to be rolled back.
Results show that there is no single perfect synchronization algorithm for this model: depending on configuration, either synchronization protocol might be better.

### 2016-06-29 publications/Journal/sequential-performance.tex

\subsubsection{Queue}
\label{4-seq-Queue}
% explain what we did: fixed number of models (400, 600 & 800), increasing depth
For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator.
A set of three tests was performed, where each test has the same number of models but an increasing depth.
The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
Since dxex symbolically flattens the model, there is no performance hit when the depth is increased.
The overhead of running the directconnect algorithm is one time only and negligible when the end time of the simulation is sufficiently large.
Adevs on the other hand does suffer from the increased depth.
With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for sequential simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}

\subsubsection{Interconnect}
\label{4-seq-Interconnect}
In the Interconnect model, we increase the number of atomic models, thus quadratically increasing the number of couplings and the number of external transitions.
As can be seen in Figure~\ref{fig:Interconnect_benchmark}, adevs outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type.
Copying and deallocation semantics and tracing are solved by the kernels at a runtime cost in simulations where event frequency is very high.
Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
We refer the interested reader to the dxex \hyperref{https://bitbucket.org/bcardoen/devs-ex-machina}{}{repo}{repository} for the profiling call graphs for the different benchmarks.

>>>>

\subsubsection{Queue}
\label{4-seq-Queue}
% explain what we did: fixed number of models (400, 600 & 800), increasing depth
For the first benchmark, we tested the effect of hierarchical complexity of the model in the performance of the simulator.
A set of three tests was performed, where each test has the same number of models but an increasing depth.
The results can be seen in Figure~\ref{fig:Queue_benchmark_seq}.
Since dxex performs direct connection on the model, there is no performance hit when the depth is increased.
Direct connection only needs to be done at initialization, so it is a one time cost that is negligible when simulation termination end time is sufficiently large.
Adevs, on the other hand, suffers from the increased depth, even though some similar (but not identical) optimization to event passing was made~\cite{adevs_opt}.
With every new hierarchical layer, routing an event from one atomic model to the next becomes more expensive, resulting in an increase in runtime.
\begin{figure}
	\center
	\includegraphics[width=\columnwidth]{fig/queue_fixed_sequential.eps}
	\caption{Queue benchmark results for sequential simulation.}
	\label{fig:Queue_benchmark_seq}
\end{figure}

\subsubsection{Interconnect}
\label{4-seq-Interconnect}
In the Interconnect model, we increase the number of atomic models, quadratically increasing the number of couplings and the number of external transitions.
As shown in Figure~\ref{fig:Interconnect_benchmark}, adevs now outperforms dxex by a fair margin.
Analysis showed that this is caused by the high amount of events: event creation is much slower in dxex than it is in adevs, despite dxex's use of memory pools.
To shield the user from threading and deallocation concerns dxex provides an event superclass from which the user can derive to create a specialized event type.
Copying, deallocation, and tracing are done at runtime, adding significant overhead when events happen frequently.
Profiling the benchmarks clearly shows the increasing cost of output generation and deallocation as the determining factor in the gap in performance.
